---
title: "Illustrative Examples"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    toc: true
    theme: simple
---

```{r setup, message = FALSE, warning = FALSE, comment = NA}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA,
                      fig.width = 6.25, fig.height = 5)

library(tidyverse)
library(Hmisc)
library(ggsci)
library(SpiecEasi)
library(ANCOMBC)
library(mia)
library(ggpubr)

folder_path = "custom_functions"
r_files = list.files(path = folder_path, pattern = "\\.R$", full.names = TRUE)
for (file in r_files) {
  source(file)
}
```

# Motivation: The effects of compositionality {.tabset}

## Data

n = 100 and d = 20. 4 pairs of features are correlated.

```{r}
set.seed(123)
n = 100
d = 20
cor_pairs = 4
mu = 10:14
da_prop = 0.1
sample_name = paste0("s", seq_len(n))
feature_name = paste0("f", seq_len(d))

# Simulate the true metabolite data
sim  =  sim_data(n = n, d = d, cor_pairs = cor_pairs, mu = mu, da_prop = da_prop)
y = sim$y
x = sim$x

# The log data
log_y = log(as.matrix(y))

# Add sources of biases
log_sample_bias = log(runif(n, 1e-3, 1e-1))
log_feature_bias = log(runif(d, 1e-1, 1))
log_data = sweep(log_y, 1, log_sample_bias, `+`)
log_data = sweep(log_data, 2, log_feature_bias, `+`)
data = exp(log_data)
```

## True Pearson correlation coefficients

```{r}
res_pearson = rcorr(log_y, type = "pearson")
pearson_values = res_pearson$r
pearson_p = res_pearson$P
pearson_q = matrix_p_adjust(pearson_p)

mat_values = p_filter(pearson_values, pearson_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_true = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Pearson correlation coefficients on the observed data

The standard Pearson correlation coefficients do not account for compositionality, which leads to inflated false positives.

```{r}
res_pearson = rcorr(log(data), type = "pearson")
pearson_values = res_pearson$r
pearson_p = res_pearson$P
pearson_q = matrix_p_adjust(pearson_p)

mat_values = p_filter(pearson_values, pearson_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_pearson = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Variance of log-ratio (VLR)

However, the VLR demonstrates robustness in the presence of compositional effects, allowing for the direct computation of VLRs.

The permutation test for deriving p-values has been modified to utilize normalized VLRs.

```{r}
outputs = vlr_test(data = data, meta = x, num_perm = 1000, p_adj_method = "BH")
estimate = outputs$estimate
p_value = outputs$p_value
q_value = outputs$q_value
mat_values = p_filter(estimate, q_value, max_p = 0.05, impute_value = 0)

df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_vlr = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "grey80", space = "Lab", name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SECOM

```{r}
feature_table = t(data)
assays = SimpleList(counts = feature_table)
smd = DataFrame(data.frame(sampleid = rownames(data)))

tse = TreeSummarizedExperiment(assays = assays, colData = smd)
res_linear = secom_linear(data = list(tse), assay_name = "counts",
                          pseudo = 0, prv_cut = 0.1, lib_cut = 0, corr_cut = 0.5, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, n_cl = 2)

secom_values = res_linear$corr_th
secom_cooccur = res_linear$mat_cooccur
feature_names = paste0("f", seq_len(d))
secom_values = secom_values[feature_names, feature_names]
secom_cooccur = secom_cooccur[feature_names, feature_names]
overlap = 10
secom_values[secom_cooccur < overlap] = 0

df_fig = data.frame(secom_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_secom = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SparCC

Compositionality-aware methods, such as SparCC and SECOM, mitigate the influence of compositional effects, consequently diminishing the occurrence of false positives.

```{r}
set.seed(123)
sparcc_values = sparcc(data)$Cor
colnames(sparcc_values) = colnames(data)
rownames(sparcc_values) = colnames(data)

# boot_res = sparccboot(data, R = 1000)
# saveRDS(boot_res, "../results/intermediate_results/illustration/boot_sparcc_composit.rds")
boot_res = read_rds("../results/intermediate_results/illustration/boot_sparcc_composit.rds")
p_res = pval.sparccboot(boot_res)
sparcc_p = diag(0, nrow = d, ncol = d)
sparcc_p[upper.tri(sparcc_p, diag = FALSE)] = p_res$pvals
sparcc_p = sparcc_p + t(sparcc_p)
colnames(sparcc_p) = colnames(data)
rownames(sparcc_p) = colnames(data)
sparcc_q = matrix_p_adjust(sparcc_p)

mat_values = p_filter(sparcc_values, sparcc_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_sparcc = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SPIEC-EASI

Other methods, such as SPIEC-EASI and mmvec, were proposed to measure the conditional probability between features.

Unlike the methods mentioned above, these approaches do not generate p-values.

```{r}
set.seed(123)
# se_mb = spiec.easi(data, method = "mb", lambda.min.ratio = 1e-2,
#                    nlambda = 20, pulsar.params = list(rep.num = 50, ncores = 4))
# saveRDS(se_mb, "../results/intermediate_results/illustration/se_composit.rds")
se_mb = read_rds("../results/intermediate_results/illustration/se_composit.rds")
se_beta = as.matrix(symBeta(getOptBeta(se_mb), mode = "maxabs"))
rownames(se_beta) = colnames(data)
colnames(se_beta) = colnames(data)

df_fig = data.frame(se_beta) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
df_fig$value[is.na(df_fig$value)] = 1
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate1_se = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

# Motivation: The effects of missing not at random (MNAR) {.tabset}

## Data

n = 100 and d = 20. 4 pairs of features are correlated.

```{r}
set.seed(123)
n = 100
d = 20
cor_pairs = 4
mu = 10:14
da_prop = 0.1
sample_name = paste0("s", seq_len(n))
feature_name = paste0("f", seq_len(d))

# Simulate the true metabolite data
sim  =  sim_data(n = n, d = d, cor_pairs = cor_pairs, mu = mu, da_prop = da_prop)
y = sim$y
x = sim$x

# The log data
log_y = log(as.matrix(y))

# Add sources of biases
log_sample_bias = log(runif(n, 1e-3, 1e-1))
log_feature_bias = log(runif(d, 1e-1, 1))
log_data = sweep(log_y, 1, log_sample_bias, `+`)
log_data = sweep(log_data, 2, log_feature_bias, `+`)
data = exp(log_data)

# Add zeros
zero_prop = 0.3
threshold = quantile(data, zero_prop, na.rm = TRUE)
data_miss = data
data_miss[data_miss < threshold] = 0
```

## Variance of log-ratios (VLR)

VLRs cannot be directly computed when 0s are present.

The most common approach to handle 0s is to add a pseudocount (e.g., 1) to all values. However, adding a pseudocount inflates the values of VLRs.

```{r}
data_pseudo = data_miss + 1
vlr_values = compute_vlr(data = data_pseudo)
norm_vlr_values = norm_vlr(vlr_values)

outputs = vlr_test(data = data_pseudo, meta = x, num_perm = 1000, p_adj_method = "BH")
estimate = outputs$estimate
p_value = outputs$p_value
q_value = outputs$q_value

mat_values = p_filter(estimate, q_value, max_p = 0.05, impute_value = 0)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate2_vlr = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "grey80", space = "Lab", name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SECOM

SECOM utilized complete data to deal with zeros, thus, with the presence of a large proportion of zeros, SECOM could suffer from loss of power.

```{r}
feature_table = t(data_miss)
assays = SimpleList(counts = feature_table)
smd = DataFrame(data.frame(sampleid = rownames(data)))

tse = TreeSummarizedExperiment(assays = assays, colData = smd)
res_linear = secom_linear(data = list(tse), assay_name = "counts",
                          pseudo = 0, prv_cut = 0.1, lib_cut = 0, corr_cut = 0.5, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, n_cl = 2)

secom_values = res_linear$corr_th
secom_cooccur = res_linear$mat_cooccur
feature_names = paste0("f", seq_len(d))
secom_values = secom_values[feature_names, feature_names]
secom_cooccur = secom_cooccur[feature_names, feature_names]
overlap = 10
secom_values[secom_cooccur < overlap] = 0

df_fig = data.frame(secom_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate2_secom = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SparCC

For methods such as SparCC, which rely on computing VLRs, the presence of 0s will lead to loss of power.

```{r}
set.seed(123)
sparcc_values = sparcc(data_miss)$Cor
colnames(sparcc_values) = colnames(data_miss)
rownames(sparcc_values) = colnames(data_miss)

# boot_res = sparccboot(data_miss, R = 1000)
# saveRDS(boot_res, "../results/intermediate_results/illustration/boot_sparcc_zero.rds")
boot_res = read_rds("../results/intermediate_results/illustration/boot_sparcc_zero.rds")
p_res = pval.sparccboot(boot_res)
sparcc_p = diag(0, nrow = d, ncol = d)
sparcc_p[upper.tri(sparcc_p, diag = FALSE)] = p_res$pvals
sparcc_p = sparcc_p + t(sparcc_p)
colnames(sparcc_p) = colnames(data_miss)
rownames(sparcc_p) = colnames(data_miss)
sparcc_q = matrix_p_adjust(sparcc_p)

mat_values = p_filter(sparcc_values, sparcc_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate2_sparcc = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## SPIEC-EASI

Methods leveraging conditional probability are generally more robust against false positives, yet they may experience a reduction in statistical power as well.

```{r}
zero_prop = 0.3
threshold = quantile(data, zero_prop, na.rm = TRUE)
data_miss = data
data_miss[data_miss < threshold] = 0

set.seed(123)
# se_mb = spiec.easi(data_miss, method = "mb", lambda.min.ratio = 1e-2,
#                    nlambda = 20, pulsar.params = list(rep.num = 50, ncores = 4))
# saveRDS(se_mb, "../results/intermediate_results/illustration/se_zero.rds")
se_mb = read_rds("../results/intermediate_results/illustration/se_zero.rds")
se_beta = as.matrix(symBeta(getOptBeta(se_mb), mode = "maxabs"))
rownames(se_beta) = colnames(data_miss)
colnames(se_beta) = colnames(data_miss)

df_fig = data.frame(se_beta) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate2_se = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

# Motivation: The effects of confounders {.tabset}

## The confounded data

n = 100 and d = 20. With no confounders, 4 pairs of features are correlated.

```{r}
set.seed(123)
n = 100
d = 20
cor_pairs = 4
mu = 10:14
da_prop = 0.1

smd = data.frame(x1 = rnorm(n = n),
                 x2 = sample(c("a", "b"), size = n, replace = TRUE))

sample_name = paste0("s", seq_len(n))
feature_name = paste0("f", seq_len(d))

# Simulate the true metabolite data
sim  =  sim_data(n = n, d = d, cor_pairs = cor_pairs, mu = mu, 
                 x = smd, cont_list = c("x1"), cat_list = c("x2"), da_prop = da_prop)
y = sim$y
x = sim$x
beta = sim$beta

# The log data
log_y = log(as.matrix(y))

# Add sources of biases
log_sample_bias = log(runif(n, 1e-3, 1e-1))
log_feature_bias = log(runif(d, 1e-1, 1))
log_data = sweep(log_y, 1, log_sample_bias, `+`)
log_data = sweep(log_data, 2, log_feature_bias, `+`)
data = exp(log_data)

# Recover the unconfounded data
log_y_unconfound = log_y - as.matrix(x) %*% t(beta)
data_unconfound = exp(log_y_unconfound)
```

## True Pearson correlation coefficients

```{r}
res_pearson = rcorr(log_y_unconfound, type = "pearson")
pearson_values = res_pearson$r
pearson_p = res_pearson$P
pearson_q = matrix_p_adjust(pearson_p)

mat_values = p_filter(pearson_values, pearson_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_true = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## True VLRs

However, the VLR demonstrates robustness in the presence of compositional effects, allowing for the direct computation of VLRs.

The permutation test for deriving p-values has been modified to utilize normalized VLRs.

```{r}
outputs = vlr_test(data = data_unconfound, meta = NULL, num_perm = 1000, p_adj_method = "BH")
estimate = outputs$estimate
p_value = outputs$p_value
q_value = outputs$q_value
mat_values = p_filter(estimate, q_value, max_p = 0.05, impute_value = 0)

df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_vlr = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "grey80", space = "Lab", name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Confounded Pearson correlation coefficients

Regardless of the presence of confounders, Pearson correlation coefficients calculated using raw data suffer from inflated false positives due to compositionality.

```{r}
res_pearson = rcorr(log_data, type = "pearson")
pearson_values = res_pearson$r
pearson_p = res_pearson$P
pearson_q = matrix_p_adjust(pearson_p)

mat_values = p_filter(pearson_values, pearson_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_pearson = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Confounded SECOM

```{r}
feature_table = t(data)
assays = SimpleList(counts = feature_table)
smd = DataFrame(data.frame(sampleid = rownames(data)))

tse = TreeSummarizedExperiment(assays = assays, colData = smd)
res_linear = secom_linear(data = list(tse), assay_name = "counts",
                          pseudo = 0, prv_cut = 0.1, lib_cut = 0, corr_cut = 0.5, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, n_cl = 2)

secom_values = res_linear$corr_th
secom_cooccur = res_linear$mat_cooccur
feature_names = paste0("f", seq_len(d))
secom_values = secom_values[feature_names, feature_names]
secom_cooccur = secom_cooccur[feature_names, feature_names]
overlap = 10
secom_values[secom_cooccur < overlap] = 0

df_fig = data.frame(secom_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_secom = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Confounded SparCC

In the presence of confounders, compositionality-aware methods like SparCC suffer from inflated false positives and loss of power.

It should be noted that some entries exhibit a reversal in their sign.

```{r}
set.seed(123)
sparcc_values = sparcc(data)$Cor
colnames(sparcc_values) = colnames(data)
rownames(sparcc_values) = colnames(data)

# boot_res = sparccboot(data, R = 1000)
# saveRDS(boot_res, "../results/intermediate_results/illustration/boot_sparcc_confound.rds")
boot_res = read_rds("../results/intermediate_results/illustration/boot_sparcc_confound.rds")
p_res = pval.sparccboot(boot_res)
sparcc_p = diag(0, nrow = d, ncol = d)
sparcc_p[upper.tri(sparcc_p, diag = FALSE)] = p_res$pvals
sparcc_p = sparcc_p + t(sparcc_p)
colnames(sparcc_p) = colnames(data)
rownames(sparcc_p) = colnames(data)
sparcc_q = matrix_p_adjust(sparcc_p)

mat_values = p_filter(sparcc_values, sparcc_q, max_p = 0.05)
df_fig = data.frame(mat_values) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_sparcc = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

## Confounded SPIEC-EASI

Conditional probability-based methods like SPIEC-EASI also experience inflated false positives and a loss of power in the presence of confounders.

```{r}
set.seed(123)
# se_mb = spiec.easi(data, method = "mb", lambda.min.ratio = 1e-2,
#                    nlambda = 20, pulsar.params = list(rep.num = 50, ncores = 4))
# saveRDS(se_mb, "../results/intermediate_results/illustration/se_confound.rds")
se_mb = read_rds("../results/intermediate_results/illustration/se_confound.rds")
se_beta = as.matrix(symBeta(getOptBeta(se_mb), mode = "maxabs"))
rownames(se_beta) = colnames(data)
colnames(se_beta) = colnames(data)

df_fig = data.frame(se_beta) %>%
  rownames_to_column("var1") %>%
  pivot_longer(cols = f1:f20, names_to = "var2", values_to = "value") %>%
  rowwise() %>%
  mutate(pair = sort(c(var1, var2)) %>% 
           paste(collapse = ",")) %>%
  group_by(pair) %>%
  distinct(pair, .keep_all = TRUE)
df_fig$var1 = factor(df_fig$var1, levels = paste0("f", seq_len(d)))
df_fig$var2 = factor(df_fig$var2, levels = paste0("f", seq_len(d)))
breaks = levels(df_fig$var1)[rep(c(TRUE, FALSE), d/2)]

fig_motivate3_se = df_fig %>%
  ggplot(aes(var1, var2, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", na.value = "grey80",
                       midpoint = 0, limit = c(-1.1, 1.1), space = "Lab", 
                       name = NULL) +
  scale_x_discrete(drop = FALSE, position = "bottom", breaks = breaks) +
  scale_y_discrete(drop = FALSE, position = "left", breaks = breaks) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.7, 0.2),
        legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                               title.position = "top", title.hjust = 0.5)) +
  coord_fixed()
```

# Outputs

```{r}
fig_motivate1 = ggarrange(fig_motivate1_true, fig_motivate1_pearson, 
                          fig_motivate1_vlr, fig_motivate1_sparcc, 
                          fig_motivate1_secom, fig_motivate1_se, 
                          nrow = 2, ncol = 3, labels = c("a", "b", "c", "d", "e", "f"))

fig_motivate2 = ggarrange(fig_motivate2_vlr, fig_motivate2_sparcc, 
                          fig_motivate2_secom, fig_motivate2_se, 
                          nrow = 2, ncol = 2, labels = c("a", "b", "c", "d"))

fig_motivate3 = ggarrange(fig_motivate3_pearson, fig_motivate3_sparcc, 
                          fig_motivate3_secom, fig_motivate3_se, 
                          nrow = 2, ncol = 2, labels = c("a", "b", "c", "d"))

ggsave(filename = "../results/figures/ext_motivate1.pdf", 
       plot = fig_motivate1, width = 12, height = 8)
ggsave(filename = "../results/figures/ext_motivate2.pdf", 
       plot = fig_motivate2, width = 9, height = 8)
ggsave(filename = "../results/figures/ext_motivate3.pdf", 
       plot = fig_motivate3, width = 9, height = 8)
```

# Session information

```{r, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```





